Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-18 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Otro poquito

\begin{abstract}
Volunteer computing is extensively used 
% (Paloma) Adverbs are always written before verbs, check this throughout the text
in science and industry, but it
still has a long way to go in order to become
% (Paloma) I would propose "a long way to go in order to become"
 a mainstream
technology in the same way the cloud technologies in general have. The
lack of % (Paloma) Add "both" here to better complete the sentence
general purpose and open source tools with built-in analytics 
% (Paloma) "built-in" is an adjective and, as such, has to be written before names
could be one of the reasons of why this is so, as well as the lack of
proofs of concept. That is why in this paper we present {\sf NodIO}, a
framework used to create client-server architectures for distributed
evolutionary algorithms, entirely written  % (Paloma) Adverb. I would also add an "and" after the comma
in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser, as well as any other client consuming its REST application
programming interface.  The architecture, along with some initial
results on the expected performance and scaling behavior when the
number of users is increased is presented. We also use
the algorithm data to find out what kind of patterns the user behavior
follow by examining the resources they devote to our experiment, and
try to fit them to some statistical distribution to create the bases
of a model of user behavior in volunteer computing experiments of this
kind that can be used by researchers to predict its performance. 
In general, our experiments prove, first, that {\sf NodIO} is a flexible and
low-overhead platform to perform distributed computation experiments
that include volunteers, and then that user behavior follows some patterns,
and, thus, are amenable to a model.
\end{abstract}

@@

There is, then, space for systems that do not make any of these
assumptions, such as the one presented in this paper, which will be
provisionally named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can
% (Paloma) be embedded in?
be embedded in any web-enabled device, including many over which 
the developers have no control.

Some time ago, our group faced the problem of diminishing funds for
buying new hardware. This was aggravated by the increasing maintenance costs and
extended downtime resulting from the continuous failures of existing
clusters.
Considering this, we leveraged 
% (Paloma) What 'something' did you leverage?
our experience in the design of web
applications, including JavaScript, since their inception and other
volunteer and
unconventional distributed evolutionary computing systems to 
design and release a new free framework that would allow anyone to
create a volunteer distributed EC experiment using cloud resources as
servers and browsers as clients.
% (Paloma) I think you should rewrite this whole sentence.
% For starters, focus in that you "leveraged" the problem through your deeply knowledge of... this and that. And "Thus/Then/Therefore", you decided... 
% (Mario) We leverage our experience in blah blah to design
% (Paloma) It may be... guess I was thinking in the noun, not in the transitive verb

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use, and
reasonable performance. But hierarchy, this is, the capability of preferentially using resources in the same organization, is a feature that could be either disregarded or included within adaptivity, which leaves us with eight desirable properties.
% (Paloma) Re-write. For example, "But is hierarchy, this is, the capability of preferentially using resources in the same organization, the one that could be either disregarded or included in adaptivity, which leaves us with eight desirable properties."
% (Paloma) And I re-wrote what I proposed yesterday.

The fulfillment of these properties could be achieved by using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first. 
% (Paloma) These should be separate sentences
However, embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed by using desktop
applications. Baratloo et al. tried in 1996 to create a system with
the aforementioned qualities called {\em Charlotte}
\cite{baratloo1996charlotte}. This system, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing systems PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system, and access to a common data store in the
server. Since the Java virtual machine is a sandbox, it is safe to the user
and as it runs compiled bytecode it also guarantees a certain level of
performance.
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. This
paper was later extended and published in a journal in 1999
\cite{baratloo1999charlotte}, but it is interesting to note that, even in the early stages of the web, the possibility of 
% (Paloma) Re-write. For example, "but it is interesting to note that, even in the early stages of the web, the possibility of..."
browser-based distributed computing was already researched.

Other systems, like JET, proposed by Soares et
al. \cite{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, allowing its use for science,
since it yields % (Paloma) Use synonyms, like "yield" or "permit". Repetitive use of "allow" is... repetitive.
 the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\cite{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. It is focused
% (Paloma) Careful with starting using plural ("Other proposals", "are") and continuig with singular. If you are referring to "SuperWeb", change "It" for "SuperWeb"
, however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers, mentioning
SPARCstations and PCs with Windows-95. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \cite{sarmenta-bayanihan} as well as the potential for
mischief of users with code in their hands
% (Paloma) I think this sentence is extremely informal
\cite{sarmenta-sabotagetolerance}. However, these early efforts by
Sarmenta once again used Java and not JavaScript, making them % (Paloma) making them
less widespread.

@@

In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library {\sf NodEO} \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
when writing code and leverages 
expertise acquired in the library itself.

@@

On the other hand, measuring the performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. But the raw material of
volunteer computing, with regard to
number of users and computation time spent by every one 
in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
system itself.

As far as we know, this paper presents one of the few experiments which % (Paloma) Change to "which" in order to not repeating "that" all the time
uses computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. The methodology used to
gather resources and the algorithms used will be described next, as
well as the results obtained with the initial setup.

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself. This
 is
why we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M}, resulting that, in fact, JavaScript can be
faster than compiled languages such as Scala and, in any case, provides
a performance that is comparable to other languages usually employed  
in evolutionary computation and numerical optimization such as Python. 
% (Paloma) Bit of a too long sentence here. Why not changing the ; for a . ?
Compiled languages such as
Java or C might be, indeed, faster, but the fact % (Paloma) Too many "fact" words. There was another one in the previous sentence.
 that a single-user
performance is higher is more than offset. This is because the number
of spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets, is several orders of magnitude
higher, which is the objective in this kind of systems.
% (Paloma) This sentence needs re-writing
% "Other compiled languages such as Java or C might be, in fact, faster, but the fact that single-user performance is higher (than what?) is over the offset. This is because the number of spontaneous volunteers we can gather at whim, including people using their mobile phones or tablets, is several orders of magnitude higher, which is the objective in this kind of systems." (?)
 JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala, or even Java, and them compile it to
JavaScript. For completness, an experiment comparing the performance of
a JavaScript implementation versus Java and Matlab is included in section
\ref{sec:w2}.

@@

Two kind of threats are possible in this environment; first, since the
API is open and it is an open source system, it is relatively easy to
sabotage the system, as indicated in \cite{domingues2007sabotage}, by
crafting a fake request which, for instance, assigns a fake fitness to
a particular chromosome
% (Paloma) Where is the second kind of threat?
. Another two different attacks are also
possible: a denial of service attack as well as penetration
testing. We have approached this threat in a pragmatic way: all
sources, client and server, for the application, are openly released. 
 We also release as open access the performed experiments \cite{DBLP:journals/corr/GuervosG15} and all data, once anonymized, as open data. This
builds trust between the user and the scientist, which indeed has an
impact on the performance of the system, since there is no need to
include cheating checks or other functions that would degrade the
performance of the system. This socio-technical approach is, indeed,
coherent with our whole approach to the study of a distributed volunteer
 system which can only be studied by approaching it from the
technical as well as the social, as in social media, angle.

In this sense, in this paper we propose the {\sf NodIO} framework, a
cloud or bare metal based volunteer evolutionary computing system
derived from % (Paloma) In order to not sounding too repetitive, could we use "derived from", for example, here?
 the {\sf NodEO} library, whose architecture has been
developed using JavaScript on the client as well as the server.
All parts of the system are free and available with a free license
from \url{https://github.com/JJ/splash-volunteer}.

@@

Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several {\em
  routes} 
  that can be called for storing and retrieving information (the CRUD cycle:
  create, request, update, and delete) from the server. 
  A JSON data format is used for the communication between 
  clients and the server. There are two kinds of information:
  {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
  GET}ing a random chromosome from it, and {\em information} related
  to the performance and state of the experiments. It also performs logging
  duties, but they are basically a very lightweight and high performance
  data storage \cite{jj:idc:lowcost}.
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found.
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with a random
  population, then it sends, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request), and requests a random
  individual from the server (via a {\tt GET} request). % (Paloma) Re-writing because there are 3 steps (island creation -> PUT -> GET)
\end{enumerate}

@@

The researcher only has to change a function, which can be written in
JavaScript or in any other language with cross-compilation to
JavaScript \cite{web:compilersjs} to solve a different
problem. In this case the classical Trap function \cite{Ackley1987} has been
used as well as the Rastrigin's floating-point optimization
problem. JavaScript is a functional language and declaring a different 
function and handing at the creation of the algorithm object, called
{\tt Classic} is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):
\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in Node.js and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a non-blocking single thread
  allows the service of many requests. In fact, a limit in the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability. 
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system or even browser requirement (as long
  as Javascript is enabled): anyone
  visiting the page, even from mobile devices, can load the algorithm.
  In case the browser does not support HTML5 Web workers (e.g., 
  Windows Phone or some version of Android) or their use
  has been disabled, a basic version of {\sf NodIO} can also be used.
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through the autonomous
  operation of every individual island with out any synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server. This architecture 
  has been tested in other high demand systems with success, but 
  in the case of {\sf NodIO}, additional experiments are need to asses 
  the scalability of the system. 
\item Since the algorithm runs in the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources that the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which means that operation is totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL and available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is better than what you would obtain in
  your own desktop system. If this is not the case, it is a pointless
  academic exercise.
\end{itemize}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{baseline-times.png}
\caption{Comparison of times to solution (labeled as {\bf times}) for the baseline system, using a {\tt
    Node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
% Plotted with: ~/Code/splash-volunteer/data/plot-baseline.R
\label{fig:baseline}
\end{figure}
First, we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using {\sf NodEO}
\cite{nodeo2014}, the basic JavaScript library we have used as base to
build {\sf NodIO}, the framework. This experiment tries to 
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal to 512. The algorithm in a
single island was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with a success rate upgraded to 100\% and an {\em average} duration of $3.46$
seconds. Results for these two experiments
are shown in Figure \ref{fig:baseline}. The baseline hardware system has these
characteristics: 
\begin{verbatim}
Linux penny 3.13.0-34-generic #60-Ubuntu SMP Wed Aug 13 15:45:27 UTC
2014 x86_64 x86_64 x86_64 GNU/Linux
\end{verbatim}
with an 8-core {\tt Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz}.

@@

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even it the user brings forward another tabbed
window.
\item Several evolutionary algorithms can be executed in parallel in a single web
page. 
\item Many implementations of the Web Worker API can use multiple-core CPUs for
the execution of worker scripts, in particular the execution in multiple-core CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each {\sf NodEO} instance can be restarted independently.
\end{itemize}

>>>>

\begin{abstract}
Volunteer computing is extensively used 
% (Paloma) Adverbs are always written before verbs, check this throughout the text
in science and industry, but it
still has a long way to go in order to become
% (Paloma) I would propose "a long way to go in order to become"
 a mainstream
technology in the same way the cloud technologies in general have. The
lack of % (Paloma) Add "both" here to better complete the sentence
general purpose and open source tools with built-in analytics 
% (Paloma) "built-in" is an adjective and, as such, has to be written before names
could be one of the reasons of why this is so, as well as the lack of
proofs of concept. That is why in this paper we present {\sf NodIO}, a
framework used to create client-server architectures for distributed
evolutionary algorithms, entirely written  % (Paloma) Adverb. I would also add an "and" after the comma
in JavaScript, which needs
minimal server support to work and is able to support persistent,
asynchronous, distributed evolutionary algorithms running in the
browser, as well as any other client consuming its REST application
programming interface.  The architecture, along with some initial
results on the expected performance and scaling behavior when the
number of users is increased is presented. We also use
the algorithm data to find out what kind of patterns the user behavior
follow by examining the resources they devote to our experiment, and
try to fit them to some statistical distribution to create the bases
of a model of user behavior in volunteer computing experiments of this
kind that can be used by researchers to predict its performance. 
In general, our experiments prove, first, that NodIO is a flexible and
low-overhead platform to perform distributed computation experiments
that include volunteers, and then that user behavior follows some patterns,
and, thus, are amenable to a model.
\end{abstract}

@@

There is, then, space for systems that do not make any of these
assumptions, such as the one presented in this paper, which will be
provisionally named {\sf NodIO}. {\sf NodIO} is basically
% (Paloma) Is the "first" word really needed? 
% (Mario) I guess is more related to fundamentally or primarily but "basically" sounds better
a two tier client/server 
application, in which clients can run on a browser. 
Therefore, they can
% (Paloma) be embedded in?
be embedded in any web-enabled device, including many over which 
developers have no control.

Some time ago, our group faced the problem of diminishing funds for
buying new hardware. This was aggravated by the increasing maintenance costs and
extended downtime resulting from the continuous failures of existing
clusters.
Considering this, we leveraged 
% (Paloma) What 'something' did you leverage?
our experience in the design of web
applications, including JavaScript, since their inception and other
volunteer and
unconventional distributed evolutionary computing systems to 
design and release a new free framework that would allow anyone to
create a volunteer distributed EC experiment using cloud resources as
servers and browsers as clients.
% (Paloma) I think you should rewrite this whole sentence.
% For starters, focus in that you "leveraged" the problem through your deeply knowledge of... this and that. And "Thus/Then/Therefore", you decided... 
% (Mario) We leverage our experience in blah blah to design

@@

The web used as a resource for distributed computing has a
long history, and almost since the beginning it was used for
evolutionary computation even before JavaScript was
introduced as a browser-based scripting language in 1997. Before that
time other technologies,
 including Flash animations and VBScript, ActiveX or Java applets were
used; for instance, Java was pointed out in \cite{soares1998get} as a
``language for the
internet'', providing some advantages such as multi-architecture compatibility or
security mechanisms, and was proposed in \cite{chandy1996world} as the
basis for the creation of a world-wide peer to peer system using compiled Java
programs. The ATLAS system
\cite{Baldeschwieler:1996:TIG:504450.504482} also used Java together
with another language called Cilk; all these papers indicated that
the introduction of an architecture-independent system such as the
Java virtual machine opened the doors for ultrascale, Internet-wide,
metacomputers. This paper also establishes the desirable features of
any {\em global computing} infrastructure, that can also be applied to
distributed evolutionary computation experiments that rely on
volunteer computing: scalability, heterogeneity, fault tolerance,
adaptive parallelism, safety, anonymity, hierarchy, ease of use and
reasonable performance. Of all these, probably {\em hierarchy}, that
is, the capability of using preferentially resources in the same
organization could be either disregarded or included in adaptivity,
leaving us with eight desirable properties.
% (Paloma) Re-write. For example, "But is hierarchy, this is, the capability of preferentially using resources in the same organization, the one that could be either disregarded or included in adaptivity, which leaves us with eight desirable properties."

The fulfillment of these properties could be achieved by using any of the
languages and associated technologies, mentioned above, provided that the user had
installed the required virtual machine plugin first, 
% (Paloma) These should be separate sentences
but embedding a
distributed computation client in the browser has the advantage of
providing a single and universal way of accessing resources via an
URI, something that is not guaranteed by using desktop
applications. Baratloo et al. tried in 1996 to create a system with
the aforementioned qualities called {\em Charlotte}
\cite{baratloo1996charlotte}. This system, which implemented a
distributed shared memory, was inspired by a previous work
on parallel computing systems PVM and MPI and solved most of the issues by
using the Java virtual machine embedded in the browser,
which provided
security to the user, uniformity of computing environment
independently of the operating system, and access to a common data store in the
server. Since the Java virtual machine is a sandbox, it is safe to the user
and as it runs compiled bytecode it also guarantees a certain level of
performance.
Other issues such as scheduling, which are related to adaptive 
parallelism, were solved on the server. This
paper was later extended and published in a journal in 1999
\cite{baratloo1999charlotte}, but it is interesting to note that,
even in the early stages of the web, the possibility of
% (Paloma) Re-write. For example, "but it is interesting to note that, even in the early stages of the web, the possibility of..."
browser-based distributed computing was already researched.

Other systems, like JET, proposed by Soares et
al. \cite{soares1998get}, also used Java. JET supports
the execution of parallel applications over the Internet and features
a comprehensive statistics collection subsystem, enabling its use for science, %(Mario) Was easier to change "allowing"
since it allows % (Paloma) Use synonyms, like "yield" or "permit" 
the collection of algorithm analytics by the
experiment designer, unlike the other systems mentioned
above. Other proposals contemporary to JET, such as {\em SuperWeb}
\cite{alexandrov1997superweb} are roughly a translation of classical
distribution techniques, with overly complicated architectures that include
brokers for resource discovery. SuperWeb is focused
% (Paloma) Careful with starting using plural ("Other proposals", "are") and continuig with singular. If you are referring to "SuperWeb", change "It" for "SuperWeb"
, however, in solving the
issues involved in browser-based computing and also in making a system
that is able to work across different types of computers, mentioning
SPARCstations and PCs with Windows-95. Interestingly enough, this
paper incorporates a discussion on the economic model, paving the way
for the for-profit distributed computing systems that arose later on. 
On the other hand, the potential for volunteer computing using
browsers was realized
later on \cite{sarmenta-bayanihan} as well as the potential of
user misconduct when having access to the source code.
% (Paloma) I think this sentence is extremely informal
\cite{sarmenta-sabotagetolerance}. However, these early efforts by
Sarmenta once again used Java and not JavaScript, making this effort % (Paloma) making them
less widespread.

@@

In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library NodEO \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
when writing code and leverages 
the expertise acquired
% (Paloma) Not sure if "acquired" should be written before "expertise"... sounds better to me, anyways
 in the library itself.

@@

On the other hand, measuring the performance of the resulting metacomputer
involves understanding the dynamics of this kind of systems. Initial
work was done for peer to peer systems by Stutzbach et
al. \cite{stutzbach2006understanding} and extended to volunteer
computing by Laredo et al. \cite{churn08,laredo2008rcp}. 
%But the raw material of           
%volunteer computing, % (Paloma) Whith regard to?  Maybe it was colloquial/analogy
But some of the essential metrics in volunteer computing like the
number of users or the time spent by every one in the
computation in browser-based volunteer computing experiments, have
only been studied in a limited way in 
\cite{DBLP:journals/gpem/LaredoBGVAGF14} on the basis of a single
run. Studies using volunteer computing platforms such as SETI@home
\cite{javadi2009mining} found out that the Weibull, log-normal and
Gamma distribution 
modeled quite well the availability of resources in several clusters
of that framework; the shape of those distributions is a skewed bell
with more resources in the {\em low} areas than in the high areas:
there are many users that give a small amount of cycles, while there
are just a few that give many cycles. This is in concordance with the
results obtained in \cite{agajaj}. Although SETI@home works
at a larger scale in both quantity of volunteers and the scope of projects, 
our initial tests indicate that
this kind of distribution might be the most appropriate to describe
user participation in several different metrics. However, in this
paper we will perform different experiments in different setups and
will try to come up with a more precise model. Next we describe the
system itself.

As far as we know, this paper presents one of the few experiments using % (Paloma) Change to "which" in order to not repeating "that" all the time
computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. 
%The methodology used
% (Paloma) I would say "the used methodology"
The algorithms used, as well as the methodology 
for gathering resources will be described next, 
together with the results obtained in this initial setup.

@@

In general, a distributed volunteer-based evolutionary computation
system based on the browser is simply a client-server system
whose client is, or it can be, embedded in the browser via
JavaScript. Since JavaScript is  the only language that is present
across all browsers, the choice was quite clear. Even so, there might
be some issues with the performance of the language itself; which % (Paloma) I recommend the use of "which" here
 is
why we have made a comparison between JavaScript and other languages
\cite{2015arXiv151101088M}. The results of this work show that JavaScript can be
faster than compiled languages such as Scala and, in any case, it provides
a performance that is comparable to other languages usually employed % (Paloma) Adverb
in evolutionary computation and numerical optimization such as Python. 
% (Paloma) Bit of too long sentence here. Why not changing the ; for a . ?
Compiled languages such as
Java or C might be faster when % (Paloma) Too many "fact" words. There was another one in the previous sentence.
considering the performance of a single-user, 
but this is more than offset by the computational power of
the spontaneous volunteers we can gather at whim, including people
using their mobile phones or tablets. Together, the performance is several orders of magnitude
higher, which is the objective in this kind of systems.
% (Paloma) This sentence needs re-writing
% "Other compiled languages such as Java or C might be, in fact, faster, but the fact that single-user performance is higher (than what?) is over the offset. This is because the number of spontaneous volunteers we can gather at whim, including people using their mobile phones or tablets, is several orders of magnitude higher, which is the objective in this kind of systems." (?)
 JavaScript is
quite a popular language nowadays, providing compilers from a number
of languages, many of them in their family (like CoffeeScript), but
many other outside, so researchers can, in fact, write their fitness
function in Python, Erlang, Scala, or even Java, and them compile it to
JavaScript. For completness, an experiment comparing the performance of
a JavaScript implementation versus Java and Matlab is included in section
\ref{sec:w2}.

@@

Two kind of threats are possible in this environment; first, since the
API is open and it is an open source system, it is relatively easy to
sabotage the system, as indicated in \cite{domingues2007sabotage}, by
crafting a fake request which, for instance, assigns a fake fitness to
a particular chromosome
% (Paloma) Where is the second kind of threat?
. Another two different attacks are also
possible: a denial of service attack as well as penetration
testing. We have approached this threat in a pragmatic way: all
sources, client and server, for the application, are released
openly. % (Paloma) Adverb
 We also release as open access experiments performed
\cite{DBLP:journals/corr/GuervosG15} and all data is also, after
cleaning it from personal information, released % (Paloma) Not sure if this "released" should be left and the first one in the sentence, erased...
 as open data. This
builds trust between the user and the scientist, which indeed has an
impact on the performance of the system, since there is no need to
include cheating checks or other functions that would degrade the
performance of the system. This socio-technical approach is, indeed,
coherent with our whole approach to the study of a distributed volunteer
 system which can only be studied by approaching it from the
technical as well as the social, as in social media, angle.

In this sense, in this paper we propose the {\sf NodIO} framework, a
cloud or bare metal based volunteer evolutionary computing system
based on % (Paloma) In order to not sounding too repetitive, could we use "derived from", for example, here?
 the {\sf NodEO} library, whose architecture has been
developed using JavaScript on the client as well as the server.
All parts of the system are free and available with a free license
from \url{https://github.com/JJ/splash-volunteer}.

@@

Thus, {\sf NodIO} architecture has two tiers:\begin{enumerate}
\item A REST server, that is, a server that includes several {\em
  routes} 
  that can be called for storing and retrieving information (the CRUD cycle:
  create, request, update, and delete) from the server. 
  A JSON data format is used for the communication between 
  clients and the server. There are two kinds of information:
  {\em problem} based, that is, related to the
  evolutionary algorithm such as {\tt PUT}ing a chromosome in or {\tt
  GET}ing a random chromosome from it, and {\em information} related
  to the performance and state of the experiments. It also performs logging
  duties, but is basically a very lightweight and high performance
  data store \cite{jj:idc:lowcost}.
  The server has the capability to
  run a single experiment, storing the chromosomes in a data structure
  that is reset when the solution is found.
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays graphs and some
  additional links and information on the experiment. This code runs
  an evolutionary algorithm {\em island} starting with a random
  population and then sending, every 100 generations, the best individual
  back to the server (via a {\tt PUT} request) and requesting a random
  individual from the server (via a {\tt GET} request).
\end{enumerate}

@@

The researcher only has to change a function, which can be written in
JavaScript or in any other language with cross-compilation to
JavaScript \cite{web:compilersjs} to solve a different
problem. In this case the classical Trap function \cite{Ackley1987} has been
used as well as the Rastrigin's floating-point optimization
problem. JavaScript is a functional language and declaring a different 
function and handing at the creation of the algorithm object, called
{\tt Classic} is the only change needed to work with a different
problem. Let us see how this system addresses the different challenges
outlined in the State of the Art (Section \ref{sec:soa}):
\begin{itemize}
\item {\em Scalability} is provided via the use of a lightweight and
  high-performance, single-threaded, server based in Node.js and
  Express.js. Although this single server is a bottleneck since it
  will eventually saturate, the fact that it runs as a non-blocking single thread
  allows the service of many requests. In fact, a limit in the
  number of simultaneous requests will be reached, but so far it has
  not been found, unlike what we found in our previous systems, DCoR,
  \cite{gecco07:workshop:dcor}, which had a low scalability. 
\item The system is {\em heterogeneous} since it does not need any
  performance, operating system or even browser requirement (as long
  as Javascript is enabled): anyone
  visiting the page, even from mobile devices, can load the algorithm.
  In case the browser does not support HTML5 Web workers (e.g., 
  Windows Phone or some version of Android) or their use
  has been disabled, a basic version of NodIO can also be used.
\item {\em Fault tolerance} is always an issue, and in this case, the
  single point of failure would be the server: the system, as such,
  would break down if the server fails. However, the individual
  islands in every browser would continue running, and having access
  to just one of them would allow the local algorithm to proceed. In
  fact, the island does not need the server to run: it runs locally if
  needed, with the only exception that it is obviously unable to
  communicate with the rest of the islands.
\item {\em Adaptiveness} is achieved simply through the autonomous
  operation of every individual island with out any synchronization
  mechanism. The islands in the system are, in fact, unaware of each
  other, communicating only through the server. This architecture 
  has been tested in other high demand systems with success, but 
  in the case of NodIO, additional experiments are need to asses 
  the scalability of the system. 
\item Since the algorithm runs in the browser, {\em safety} is
  achieved through its sandbox mechanisms. The user is thus assured
  that there is no unsafe access either to their local files or even
  to more resources that the browser should be allotted.
\item Running the algorithm is just a matter of loading the page,
  which means that operation is totally {\em anonymous}. For the same
  reason, {\em ease of use} is optimal, being as easy as simply
  clicking on an URL and available to anyone with access to a browser.
\item {\em Reasonable performance} is not ensured. In fact, we should
  make sure that there is a reasonable amount of clients over which
  the performance achieved is better than what you would obtain in
  your own desktop system. If this is not the case, it is a pointless
  academic exercise.
\end{itemize}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{baseline-times.png}
\caption{Comparison of times to solution (labeled as {\bf times}) for the baseline system, using a {\tt
    Node.js} client, for different population sizes. Only
runs in which solution has been found are used for this graph.}
% Plotted with: ~/Code/splash-volunteer/data/plot-baseline.R
\label{fig:baseline}
\end{figure}
First, we establish the baseline performance by running the
evolutionary algorithm in a desktop client written using NodEO
\cite{nodeo2014}, the basic JavaScript library we have used as base to
build {\sf NodIO}, the framework. This experiment tries to 
find the solution to the 40-trap function with parameters $l=4, a=1,
b=2, z=3$ and a population size equal to 512. The algorithm in a
single island was run until the solution (a string with all
ones) was found or five million evaluations had been performed. It
took around a minute, on average, that is $68.9694$ seconds, to
perform the fifty runs. In this experiment only 33, that is, 66\% were
successful. The experiments were repeated for population $p=1024$,
with a success rate upgraded to 100\% and an {\em average} duration of $3.46$
seconds. Results for these two experiments
are shown in Figure \ref{fig:baseline}. The baseline hardware system has these
characteristics: 
\begin{verbatim}
Linux penny 3.13.0-34-generic #60-Ubuntu SMP Wed Aug 13 15:45:27 UTC
2014 x86_64 x86_64 x86_64 GNU/Linux
\end{verbatim}
with an 8-core {\tt Intel(R) Core(TM) i7-4770 CPU @ 3.40GHz}.

@@

\begin{itemize}
\item If the browser uses a tabbed document interface, the worker script
keeps running in the background even it the user brings forward another tabbed
window.
\item Several evolutionary algorithms can be executed in parallel in a single web
page. 
\item Many implementations of the Web Worker API can use multiple-core CPUs for
the execution of worker scripts, in particular the execution in multiple-core CPUs
was tested in desktop versions of Chrome, FireFox and Safari.
\item Each NodEO instance can be restarted independently.
\end{itemize}
